Concurrent Programming with Harmony

### 2020-08-27 book/paper.tex
Latest

The most common use for locks is in building concurrent data structures.
By way of example, we will demonstrate how to build a concurrent queue.
The queue will have the following API:
\begin{itemize}
\item $q = \mathtt{Qnew}()$: allocate a new queue;
\item $Qenqueue(q, v)$: add $v$ to the tail of the queue;
\item $r = Qdequeue(q)$: returns $r = ()$ if $q$ is empty or $r = (v,)$ (a singleton tuple)
if $v$ was at the head of the queue.
\end{itemize}
See \autoref{fig:queuetest} for a simple test program that uses the queue.

@@

The implementation in \autoref{fig:queueMS}~\cite{MS98} is similar but uses
separate locks for the head and tail, allowing an enqueue and a dequeue operation
to proceed concurrently.  To avoid contention between the head and the tail that
requires both locks, the queue uses a dummy node at the head of the linked list.
Except of the initial state of the queue, the dummy node is the last node that
was dequeued.  Note that neither the \texttt{head} nor \texttt{tail} pointer are
ever \textbf{None}.

@@

\section*{Exercises}
\begin{problems}
\item Assume that you have only binary semaphores.
Implement counting semaphores using the split binary semaphores technique.
\item With reader/writer locks,
it can be useful for a process to \emph{downgrade} its write lock
to a read lock when its done writing but not done reading.  Add
a \texttt{downgrade} method to the code in \autoref{fig:RWsplitsema}.
How easy or hard would it be to do the same for the code in
\autoref{fig:rw2lock}?
Note, you cannot simply release the
write lock and then obtain the read lock, because the process needs
to remain in the critical section.
\item \label{ex:onelane} Implement a solution to the producer/consumer problem
(\autoref{fig:boundedbuffer}) using split binary semaphores.
\item \label{ex:onelane} Cornell's campus features some one-lane bridges.
On a one-lane bridge,
cars can only go in one direction at a time. Consider northbound
and southbound cars wanting to cross a one-lane bridge.
The bridge allows arbitrary many cars, as long as they're going in the
same direction.
Implement a lock that observes this requirement using SBS.  Write methods
\texttt{nb\_enter()} that a car must invoke before going northbound on
the bridge and \texttt{nb\_leave()} that the car must invoke after leaving
the bridge.  Similarly write \texttt{sb\_enter()} and \texttt{sb\_leave()}
for southbound cars.
\item Extend the solution to \autoref{ex:onelane} by implementing the
requirement that at most 3 cars are allowed on the bridge.

>>>>

The most common use for locks is in building concurrent data structures.
By way of example, we will demonstrate how to build a concurrent queue.
The queue will have the following API:
\begin{itemize}
\item $q = \mathtt{Qnew}()$: allocate a new queue;
\item $\mathtt{Qenqueue}(q, v)$: add $v$ to the tail of the queue;
\item $r = \mathtt{Qdequeue}(q)$: returns $r = ()$ if $q$ is empty or $r = (v,)$ (a singleton tuple)
if $v$ was at the head of the queue.
\end{itemize}
See \autoref{fig:queuetest} for a simple test program that uses the queue.

@@

The implementation in \autoref{fig:queueMS}~\cite{MS98} is similar but uses
separate locks for the head and tail, allowing an enqueue and a dequeue operation
to proceed concurrently.  To avoid contention between the head and the tail that
requires both locks, the queue uses a dummy node at the head of the linked list.
Except for the initial state of the queue, the dummy node is the last node that
was dequeued.  Note that neither the \texttt{head} nor \texttt{tail} pointer are
ever \textbf{None}.

@@

\section*{Exercises}
\begin{problems}
\item Assume that you have only binary semaphores.
Implement counting semaphores using the split binary semaphores technique.
\item With reader/writer locks,
concurrency can be improved if a process \emph{downgrades} its write lock
to a read lock when its done writing but not done reading.  Add
a \texttt{downgrade} method to the code in \autoref{fig:RWsplitsema}.
How easy or hard would it be to do the same for the code in
\autoref{fig:rw2lock}?
Note, you cannot simply release the
write lock and then obtain the read lock, because the process needs
to remain in the critical section.
\item \label{ex:onelane} Implement a solution to the producer/consumer problem
(\autoref{fig:boundedbuffer}) using split binary semaphores.
\item \label{ex:onelane} Cornell's campus features some one-lane bridges.
On a one-lane bridge,
cars can only go in one direction at a time. Consider northbound
and southbound cars wanting to cross a one-lane bridge.
The bridge allows arbitrary many cars, as long as they're going in the
same direction.
Implement a lock that observes this requirement using SBS.  Write methods
\texttt{nb\_enter()} that a car must invoke before going northbound on
the bridge and \texttt{nb\_leave()} that the car must invoke after leaving
the bridge.  Similarly write \texttt{sb\_enter()} and \texttt{sb\_leave()}
for southbound cars.
\item Extend the solution to \autoref{ex:onelane} by implementing the
requirement that at most 3 cars are allowed on the bridge.
